---
title: "Final Project"
author: "Holly Figueroa"
date: "5/22/2021"
output:
  pdf_document: default
  pdf: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning= FALSE, message=FALSE)
```
A central dataset to be used from the census bureau includes numbers of those willing to get a vaccination, once available.This table contains extra information in the first rows that will have to be skipped. Columns must also be renamed for simplicity as follows:

|Column Name   |New Name|
|---------------|---------|
|Week | week | 
|Area | state |
|Total.Individual.Population. age.18 | state_adult_pop |
|Measure.Universe | willing_sample |
|Number | total_willing |
|Percent.Margin.of.Error.... | pc_MoE_willing |
|Percent | pc_willing |

I will then add other variables of interest on the state level from the Pulse Survey by opening those tables, keeping selected columns to combine together later. I will create a data frame containing percentages of people from each state that report being willing to get a vaccine, those who expected income loss, those who already experienced income loss, those who expected eviction, and those who had delayed relieving medical care. The data files created to be combined together are as follows, all percentages are given at the state level:  

|Data Name|Contents|
|--|-----|
|vaccine_willing_percents||
|income_lost_percents|
|exp_income_loss_percents|
|exp_eviction_percents|
|election2020| percentage republican votes of total state votes|

A data frame of census survey variables relating to metro cities was separated out from the state data and set aside for potential use. I do not have election data at this level of measurement, however, so any analysis would not involve election variables. Data included from the census also has a first row including the United States as a whole. This was also taken out and set aside for potential reference. With all the variables combined the complete data set contains the following variables:

|Variable Name |Variable Meaning|
|-----|-----|
|state| state|
|pc_willing|percentage of individuals planning or willing to vaccinate once able|
|pc_exp_income_loss|percentage of individuals that anticipated a loss of income in the next 4 weeks|
|pc_income_lost|percentage of households where someone had a loss in employment income in the last 7 days|
|pc_ex_eviction|percentage of individuals that expected eviction or home foreclosure in the next two months|
|trump_percentage|percentage of votes that were won by Donald Trump out of all presidential votes cast|



```{r}
library(dplyr)
# WILLINGNESS CHANGES
# Open and clean willingness to vaccinate dataframe
orig_vaccine_df <- read.csv('final_project/vaccine_will.csv', skip=1)
head(orig_vaccine_df)
# rename columns and view
orig_vaccine_df <-orig_vaccine_df%>%
  rename(week = Week, 
         state = Area, 
         state_adult_pop = Total.Individual.Population.age.18., 
         willing_sample = Measure.Universe,
         total_willing = Number, 
         pc_MoE_willing = Percent.Margin.of.Error...., 
         pc_willing = Percent)
head(orig_vaccine_df)

# Select columns to keep and combine with others later.
vaccine_willing_percents <- orig_vaccine_df%>%
  select(state, pc_willing)
head(vaccine_willing_percents)

# EXPECTED INCOME LOSS CHANGES
# read expected loss of income due to Covid data file  
orig_exp_income_loss_df <- read.csv('final_project/exp_income_loss.csv', skip = 1)
head(orig_exp_income_loss_df)
colnames(orig_exp_income_loss_df)

# create percentages only data frames to combine later and rename columns
exp_income_loss_percents <- orig_exp_income_loss_df%>%
  select(2,7)%>%
  rename(state = Area, pc_exp_income_loss = Percent)
head(exp_income_loss_percents)

# INCOME LOST CHANGES
# read data on people with income lost due to Covid data file
orig_income_lost_df <-read.csv('final_project/income_lost.csv', skip = 1)
head(orig_income_lost_df)
colnames(orig_income_lost_df)

# create percentages only data frame to combine later and rename columns
income_lost_percents <-orig_exp_income_loss_df%>%
  select(2,7)%>%
  rename(state= Area, pc_income_lost = Percent)
head(income_lost_percents)

# EXPECTED EVICTION CHANGES
# read data file on people who anticipated eviction/foreclosure
orig_exp_eviction_df <-read.csv('final_project/eviction_likely.csv', skip = 1)
head(orig_exp_eviction_df)
colnames(orig_exp_eviction_df)

# create percentages only data frame to combine later and rename columns
exp_eviction_percents <- orig_exp_eviction_df%>%
  select(2,7)%>%
  rename(state = Area, pc_exp_eviction = Percent)
head(exp_eviction_percents)

# DELAYED MEDICAL CARE CHANGES
# read data file on people who delayed receiving medical care due to Covid
orig_delayed_med_df <- read.csv('final_project/delayed_med.csv', skip = 1)
head(orig_delayed_med_df)
colnames(orig_delayed_med_df)

#create percentages only data frame to combine later and rename columns
delayed_med_percents <- orig_delayed_med_df%>%
  select(2,7)%>%
  rename(state = Area, pc_delayed_med = Percent)
head(delayed_med_percents)

# Combine data frames into one, check, and tidy
my_data <-cbind(vaccine_willing_percents,
                exp_income_loss_percents,
                income_lost_percents,
                exp_eviction_percents,
                delayed_med_percents)
colnames(my_data)

# Take out duplicate state columns
my_data<-my_data%>%
  select(-3, -5, -7, -9 )
head(my_data)

# Separate out city metro data into separate file
library(stringr)
metro_data <- my_data%>%
  filter(str_detect(state, "Metro"))%>%
  rename(location = state)
head(metro_data)

# Slice out metro data from my_data
nrow(my_data)
my_data <- slice(my_data,c(1:52))

# Separate out United States level of observation
us_census_data <- my_data%>%
  filter(state == "United States")
head(us_census_data)

# Slice out United States level of observations so only data on 51 states remains
my_data <- slice(my_data, c(2:52))
nrow(my_data)
  


```

```{r}
# ELECTION DATA CHANGES

# PRESIDENTIAL DATA
election2020_state_and_county <- read.csv('final_project/president_county_candidate.csv')
head(election2020_state_and_county)

# Get total pres votes by state
election2020<-election2020_state_and_county%>%
  group_by(state)%>%
  summarise_at(vars(total_votes), list(total_votes = sum))

# get republican pres votes by state
rep_votes <- election2020_state_and_county%>%
  filter(candidate == "Donald Trump")%>%
  group_by(state)%>%
  summarise_at(vars(total_votes), list(trump_votes = sum))%>%
  select(trump_votes)

head(rep_votes)

# Combine columns: state,total presidential votes, and total presidential votes 
election2020 <- cbind(election2020,rep_votes)

# Create percentage republican presidential votes column
election2020$trump_percentage <- (election2020$trump_votes / election2020$total_votes) * 100

#Rename column to specify presidential total votes
election2020 <- election2020%>%
  rename(total_pres_votes = total_votes)
head(election2020)

# Explore table
summary(election2020)

```

```{r}
# COMBINE ALL VARIABLES AT THE STATE LEVEL INTO ONE DATA FRAME

nrow(my_data)
nrow(election2020)
combined_data <- merge(x = my_data, y = election2020)
head(combined_data)

combined_data <- combined_data%>%
  select(-trump_votes,-total_pres_votes)
# Head final table
head(combined_data)

#Rename columns to avoid issues calling the data by shared starting text
combined_data <- combined_data%>%
  rename(wiling_pc = pc_willing, 
                        exp_income_loss_pc = pc_exp_income_loss,
                        income_lost_pc = pc_income_lost,
                        exp_eviction_pc = pc_exp_eviction,
                        delayed_med_pc = pc_delayed_med)

head(combined_data)

```

